Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2023-05-06 analysis/paper/paper.qmd
Edited based on Fitzhugh's comments

Stemmed points (Sumbejjirugae in Korean) are projectile points made on an elongated blade-like flake or blade with two parallel facets and a single or two ridges that converge to form an inverted "Y" [@pratt2020circum] (@fig-sp-hand-made). Slight retouch is typically performed on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a wooden shaft. Elsewhere in the world these types of artifacts are often called "tanged points", but we prefer "stemmed points" to distinguish them from Bronze Age stone projectile points known as "tanged points" in Korea [@park2022did]. Understanding the appearance of stemmed points is relevant to general questions about the direction of projectile technology, the technological transition into the Late Paleolithic, and relationships between Korea and adjacent regions in East Asia, such as Japan.

@@

Since the first appearance of stemmed points defines the beginning of the Korean Late Paleolithic, investigating their origins is critical to understanding the technological transition from the Early Paleolithic, modern human dispersals into the region, and claims for the existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern]. The debate around the origin of stemmed points can be summarized into two competing models [@bae2013early;@lee2016patterns] : (1) a 'heterogenic' migration [@bae2010origin], and (2) an *in situ* evolution model [@Seong_2006]. The migration model claims that the new blade industry - including stemmed points - and the earlier coarse flake tradition - including large cores, polyhedrons, choppers, and even handaxes - came from different origins. These are (1) a Northern route: Siberia, Mongolia, or other regions of northeast China, and (2) a Southern route: southern China) as the result of a continuing influx of modern human migration from two routes. The *in situ* model argues that stemmed points and other Late Paleolithic technologies, such as blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence. The difference between the two models comes from how they explain a few early sites and stemmed points made out of flakes. The *in situ* model argues that stemmed points appeared before blades, and identifies early sites such as Bonggok, Songamri, Yonghodong, and Hwadaeri as evidence that the Korean Late Paleolithic began with the emergence of stemmed points. In contrast, the migration model contends that the Late Paleolithic began with the introduction and widespread use of blade technology, similar to the traditional definition of the Late or Upper Paleolithic in Europe and the Later Stone Age in Africa. [@seong2016eastern]. Besides these two main models, there are combination of migration and trade interaction model [@bae2012nature] and complex and non-directional model [@lee2016patterns].

@@

The tip cross-sectional area (TCSA) of stone artifacts has been used as a ballistically relevant standard to probabilistically discriminate between likely weapon-tip types, such as spear-thrower (a.k.a. atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins; @lombard2022standardized; @metz2023bow; @lombard2023neanderthal; @lombard2022tip]. It is critical to note that the TCSA metric alone cannot unambiguously determine artifact function; it only suggests a best-fit ballistic probability for the points if they were hafted as weapon tips. The TCSA metric represents the part of the tool that cuts the target's hide and is related to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation]. This method was first proposed by @hughes1998getting and validated by @shea2006origins through the comparison of archaeological examples with ethnographically collected samples of known use. @shea2009impact also applied the approach to compare projectile points from Africa, the Levant, and Europe, claiming that projectile weapons first appeared in Africa.

@@

We also explored spatial patterns using 16 geographical zones based on geographical characteristics including inland, coastal areas, major rivers, islands, and major mountain ranges in addition to ecological information including temperature and precipitation [@lee2008relationship]. The sites in our sample occur in 14 of these zones: Imjin river basin (IRB), Metropolitan (MP), Central inland (CI), Kangwon coastal (KC), Choongnam coastal (CC), Southwestern inland (SWI), Upper Nagdong river basin (UNRB), WoolYoung coastal (WYC), Western Cholla (WC), Southern mountain (SM), Southeastern inland (SEI), Hyungsan Taewha coastal (HTC), Western south coastal (WSC), and Eastern south coastal (ESC). Then, we used longitude and latitude to locate individual assemblages and examined TCSA values for each zone. We acknowledge that modern eco-regional maps may not fully resemble the Pleistocene landscape in which the points were made and used; however, these are the only maps of eco-regional zones that are currently available.

@@

Our results indicate that javelin tips and stabbing spear tips are the most probable ballistic uses for stemmed points. There are a few stemmed points in our sample that may have been used as dart tips and un-poisoned arrow tips. None have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp). In their study, @lee2019were calculated TCSA values for ten stemmed points from Jingeuneul site, which were not included in this research due to the unavailability of their data. They claimed that these values are within the range of North American dart tips and arrowheads. We found a wide range of TCSA in our sample covering the range of dart tips and un-poisoned arrow tips, which are assumed to be equivalent to arrowheads. In general, the wide range of TCSA suggest that stemmed points might play diverse roles in foraging toolkits.

>>>>

Stemmed points (Sumbejjirugae in Korean) are projectile points made on an elongated blade-like flake or blade with two parallel facets and a single or two ridges that converge to form an inverted "Y" [@pratt2020circum] (@fig-sp-hand-made). Slight retouch is typically performed on the distal end to shape an acute tip and on the proximal end to make a stem, which connects to a wooden shaft. Elsewhere in the world these types of artifacts are often called "tanged points", but we prefer "stemmed points" to distinguish them from Bronze Age stone projectile points known as "tanged points" in Korea [@park2022did]. Understanding the appearance of stemmed points is relevant to general questions about the direction of projectile technology, the technological transition into the Late Paleolithic, and relationships between Korea and adjacent regions in East Asia, such as Japan.

@@

Since the first appearance of stemmed points defines the beginning of the Korean Late Paleolithic, investigating their origins is critical to understanding the technological transition from the Early Paleolithic, modern human dispersals into the region, and claims for the existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern]. The debate around the origin of stemmed points can be summarized into two competing models [@bae2013early;@lee2016patterns] : (1) a 'heterogenic' migration [@bae2010origin], and (2) an *in situ* evolution model [@Seong_2006]. The migration model claims that the new blade industry - including stemmed points - and the earlier coarse flake tradition - including large cores, polyhedrons, choppers, and even handaxes - came from different origins. These are (1) a Northern route: Siberia, Mongolia, or other regions of northeast China, and (2) a Southern route: southern China) as the result of a continuing influx of modern human migration from two routes.

The *in situ* model argues that stemmed points and other Late Paleolithic technologies, such as blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence. The difference between the two models comes from how they explain a few early sites and stemmed points made out of flakes. The *in situ* model argues that stemmed points appeared before blades, and identifies early sites such as Bonggok, Songamri, Yonghodong, and Hwadaeri as evidence that the Korean Late Paleolithic began with the emergence of stemmed points. In contrast, the migration model contends that the Late Paleolithic began with the introduction and widespread use of blade technology similar to the traditional definition of the Late or Upper Paleolithic in Europe and the Later Stone Age in Africa, without counting the stemmed points made out of non-blade materials. [@seong2016eastern]. Besides these two main models, there are combination of migration and trade interaction model [@bae2012nature] and complex and non-directional model [@lee2016patterns].

@@

The tip cross-sectional area (TCSA) of stone artifacts has been used as a ballistically relevant standard to probabilistically discriminate between likely weapon-tip types, such as spear-thrower (a.k.a. atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins; @lombard2022standardized; @metz2023bow; @lombard2023neanderthal; @lombard2022tip]. It is critical to note that the TCSA metric alone cannot unambiguously determine artifact function; it only suggests a best-fit ballistic probability for the points if they were hafted as weapon tips. The TCSA metric represents the part of the tool that cuts the target's hide and is related to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation]. This method was first proposed by @hughes1998getting and validated by @shea2006origins through the comparison of archaeological examples with ethnographically collected samples of known use. Lombard et al. demonstrated the efficacy of TCSA by analyzing Middle Stone Age points from Sibudu Cave in KwaZulu-Natal, South Africa, which are dated to approximately 35-50 ka. They compared the results obtained from experimental and use-trace studies such as organic residue analysis, macrofracture analysis and use-wear analysis with TCSA values [@lombard2021variation;@lombard2004distribution;@lombard2005evidence). @shea2009impact also applied the approach to compare projectile points from Africa, the Levant, and Europe, claiming that projectile weapons first appeared in Africa.

@@

We also explored spatial patterns using 16 geographical zones based on geographical characteristics including inland, coastal areas, major rivers, islands, and major mountain ranges in addition to ecological information including temperature and precipitation [@lee2008relationship]. The sites in our sample occur in 14 of these zones: Imjin river basin (IRB), Metropolitan (MP), Central inland (CI), Kangwon coastal (KC), Choongnam coastal (CC), Southwestern inland (SWI), Upper Nagdong river basin (UNRB), WoolYoung coastal (WYC), Western Cholla (WC), Southern mountain (SM), Southeastern inland (SEI), Hyungsan Taewha coastal (HTC), Western south coastal (WSC), and Eastern south coastal (ESC). Then, we used longitude and latitude to locate individual assemblages and examined TCSA values for each zone. We are fully acknowledged that modern eco-regional maps do not fully resemble the Pleistocene landscape in which the points were made and used. However, in the absence of detailed paleoenvironmental information, we thought it is still meaningful to adapt the modern information of vegetation and geography by assuming that some environmental factors such as  altitude, slope, drainage, hydrology, and bedrock are relatively stable over time. And those factors enable us to think that the modern vegetation patterns can be used to make reasonable references about the past.

@@

Our results indicate that javelin tips and stabbing spear tips are the most probable ballistic uses for stemmed points. There are a few stemmed points in our sample that fall within the ranges of dart tips and un-poisoned arrow tips. None have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp). In their study, @lee2019were calculated TCSA values for ten stemmed points from Jingeuneul site, which were not included in this research due to the unavailability of their data. They claimed that these values are within the range of North American dart tips and arrowheads. We found a wide range of TCSA in our sample covering the range of dart tips and un-poisoned arrow tips, which are assumed to be equivalent to arrowheads. In general, the wide range of TCSA suggest that stemmed points might play diverse roles in foraging toolkits.
